---
title: "peakPerformR: Athlete Prime Analysis Package"
author: "Jacob Berlin and Eli Vatsaas"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(peakPerformR)
library(dplyr)
library(ggplot2)
```

## Introduction


## Building Sports Data

First, let's build data for all sports using the cached data available in the package:

```{r build_data}
# Build all sports data
all_sports <- build_all_sports()
# Display summary of available sports data
summary(all_sports)
```

```{r filter_players}
players_with_sufficient_data <- all_sports %>%
  group_by(id) %>%
  summarize(num_seasons = n_distinct(season),
            player_name = first(player_name)) %>%
  filter(num_seasons >= 5) %>%
  select(player_name, id)
print(paste("Number of players with 5+ seasons:", length(players_with_sufficient_data$player_name)))
filtered_data <- all_sports %>%
  filter(id %in% players_with_sufficient_data$id)
```
## Processing Player Trajectories

Now, let's process player trajectories with specific filters:

```{r process_trajectories,message=FALSE, warnings=FALSE}
# Apply filters for player trajectories
player_trajectories <- process_player_trajectories(filtered_data)
# View the first few trajectories
head(player_trajectories$trajectories)

# Count successful models by sport
model_counts <- player_trajectories$models %>%
  left_join(filtered_data %>% select(id, league) %>% distinct(), by="id") %>%
  group_by(league) %>%
  summarize(
    total_players = n(),
    successful_models = sum(fit_success),
    success_rate = round(successful_models/total_players*100, 1)
  )

print(model_counts)
```
## Identifying Prime Years
After processing trajectories, we can identify prime years for each athlete using the identify_prime() function. This determines when an athlete's performance was above a specified threshold relative to their peak:
```{r identify-prime}
# Create data for identify_prime function
player_joins <- player_trajectories$trajectories %>%
  left_join(filtered_data %>% select(id, league, position, age, games_played), 
            by = c("id", "age"))

# Identify prime years using predicted trajectory data
player_primes_pct <- identify_prime(
  player_joins, 
  method = "predicted", 
  threshold_pct = 70,  # 70% of peak performance
  games_pct_threshold = 40  # Threshold for games played
)

# View the first few results
head(player_primes_pct)

# Count athletes by prime duration
prime_duration_counts <- player_primes_pct %>%
  mutate(prime_duration_rounded = round(prime_duration)) %>%
  group_by(prime_duration_rounded) %>%
  summarize(count = n()) %>%
  arrange(prime_duration_rounded)

print(prime_duration_counts)
```
## Clustering season performance and Creating Performance Metrics
To prepare our data for PQI calculation, we need to create a performance dataset and add metrics like tier scores and in-prime flags:
```{r cluster}
# Assuming you have a cluster_data object from clustering analysis
# For this example, we'll create a simplified version
cluster_data <- filtered_data %>%
  group_by(id) %>%
  summarize(
    cluster = sample(1:3, 1),  # Random cluster for demonstration
    .groups = "drop"
  )

# Create performance dataset
player_data <- create_player_performance_dataset(
  sports_data = filtered_data,
  prime_data = player_primes_pct,
  cluster_data = cluster_data
)

# View the structure
str(player_data[, 1:15])  # Show first 15 columns

# Examine in_prime distribution
prime_summary <- player_data %>%
  group_by(id) %>%
  summarize(
    total_seasons = n(),
    prime_seasons = sum(in_prime, na.rm = TRUE),
    prime_pct = round(prime_seasons/total_seasons*100, 1)
  ) %>%
  arrange(desc(prime_pct))

head(prime_summary)
```
## Calculating Prime Quality Index (PQI)
Now we can process the prime metrics and calculate the Prime Quality Index (PQI), which quantifies an athlete's career value based on prime performance:
```{r pqi}
# Process player primes to get aggregated metrics
processed_data <- process_player_primes(player_data)

# Calculate PQI scores
pqi <- calculate_prime_quality_index(
  processed_data,
  nfl_by_position = TRUE,  # Use position-specific normalization for NFL
  exclude_positions = c("OL"),  # Exclude certain positions
  save_output = FALSE
)

# Show top players by PQI score
top_players <- pqi %>%
  select(id, player_name, league, position, pqi_selected, prime_tier) %>%
  arrange(desc(pqi_selected)) %>%
  head(10)

print(top_players)

# Distribution of PQI tiers by league
tier_distribution <- pqi %>%
  group_by(league, prime_tier) %>%
  summarize(count = n(), .groups = "drop") %>%
  arrange(league, prime_tier)

print(tier_distribution)
```
